Inference using context-dependent models of mutation

### 2017-05-03 writeups/context-dependence.tex
begun ising testing
various adjustments

@@

\peter{from simulation: parameter estimation; model selection}

\peter{do a three-taxon tree? infer with shape data?}

@@

\subsection{TASEP}

>>>>

%%%%%%%%
\subsection{Model fit and model selection}

The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data, 
and $E_{x,y}$ for the number expected under the model.
(The expected number is the number of occurrences of $x$ in the initial sequence 
multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

@@

\textbf{To-do:}

1. Posterior coverage of parameters for Ising in 20 runs (more?).

2. With a CpG model with two 3mer motifs added, 1. fit single-base model; 2. add all 2mers; 3. add top one if significant by LR test and repeat; 4. do same for 3mers

3. Analyze human data with shape model, Bayesian.

@@

%%%%%%%%%%%
\subsection{Stepwise model selection}

In exploratory application to real data, how does one decide when to stop adading mutational motifs?
Perhaps the simplest method conceptually is to 
(a) fit a simple model,
(b) choose new motifs to add based on statistically significant residuals,
(c) repeat until no residuals are statisically significant.
To test this method,
we simulated a single sequence of $10^6$ nucleotides
for one units of time
under the model
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &    $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
        \nC\nG     &  \nT\nG     &  $0.4$ & \\
        \nC\nG     &  \nC\nA     &  $0.4$ & \\
        \nT\nC\nC  &  \nT\nT\nC  &  $0.1$ & \\
        \nA\nG\nG  &  \nA\nA\nG  &  $0.1$ &
      \end{tabular} ,
    \end{center}
with $m_{AT}=m_{TA}=0.1$, $m_{CG}=m_{GC}=0.15$, $m_{AC}=m_{TG}=m_{AG}=m_{TC}=0.08$, and $m_{CA}=m_{GT}=m_{CT}=m_{GA}=0.12$.
% Note the model is symmetric under reverse complementation.
The resulting sequences differed at 28.5\% of the sites.
We then fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
(three sites at the base with an overhang of one on each side).
The top and bottom of the table of (2,1,0) and (2,0,1) T-mer residuals of this model were as follows:
    \begin{center}
        XXX
    \end{center}
The situation clearly calls for addition of the CpG mutation tuple.
After adding these patterns into the model and re-fitting,
some length 2 residuals had statistically significant residuals,
but none nearly so large as those seen in (3,1,1) T-mer residuals:
    \begin{center}
        XXX
    \end{center}
\ldots which again clearly call for adding the appropriate 3mer mutation tuples.
After adding these in, no further (3,1,1) T-mer residuals achieved
statistical significance after a Bonferroni correction;
the final parameter values are shown in Table XXX;
parameter estimates differ from the truth by less than X\%.

\begin{table}
  \begin{center}
    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
  \end{center}
  \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
    with the above mutational motifs.
    \label{tab:cpg_results} }
\end{table}

%%%%%%%%%%%
\subsection{Bayesian inference}

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.15$, $\beta=0.3$, and $\gamma=0.15$
for 1 unit of time.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$.

%%%%%%%%%%%
\subsection{TASEP}
